Retrofitting Effect Handlers onto OCaml

### 2020-11-16 retro-concurrency.tex
edits

\begin{abstract}
Effect handlers have been gathering momentum as a mechanism for modular
	programming with user-defined effects. Effect handlers allow for non-local
	control flow mechanisms such as generators, async/await, lightweight threads
	and coroutines to be composably expressed. We present a design and evaluate a
	full-fledged efficient implementation of effect handlers for OCaml, an
	industrial-strength multi-paradigm programming language. Our implementation
	strives to maintain the backwards compatibility and existing performance
	profile of conventional OCaml code. Retrofitting effect handlers onto OCaml
	is challenging since OCaml does currently not have any non-local control flow
	mechanisms other than exceptions. Our implementation of effect handlers for
	OCaml: {\em (i)} imposes negligible overhead on code that does not use effect
	handlers; {\em (ii)} remains compatible with program analysis tools that
	inspect the stack; and {\em (iii)} is efficient for new code that makes use
	of effect handlers.
\end{abstract}

@@

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are defined by \emph{handlers} of the
effects. For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

@@

Line 4 declares an effect |DivBy0| that is parameterised by an integer,
which when performed returns the result of division-by-zero. If the denominator
is 0, then we perform the |DivBy0| effect with the numerator value (line 12) so that
the context may decide the return value. We obtain the |iOS| semantics by
wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

The need to host millions of continuations at the same time rules out the use
of a large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions (which do not have stack
overflow checks) must be performed on a separate system stack. Additionally, we
must generate DWARF stack unwind tables for stacks that may be non-contiguous.
In this work, we develop the compiler and runtime support for implementing
efficient effect handlers for OCaml that satisfy these requirements.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, pushes an exception handler frame that forwards the
exception to the innermost OCaml exception handler (|raise_exn_c| in
Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
common to use them for \emph{local} control flow.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting the GC. For every call point in the program, the OCaml
compiler emits the size of the frame and the set of all live registers in the
frame containing pointers into the heap. During a GC, the OCaml stack is walked
and the roots are marked skipping over the C frames.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end of
input, which is handled by the exception handler which closes the channels. The
|close_*| functions do nothing if the channel is already closed. The code is
written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the channels are
closed and the exception is reraised to communicate the exceptional behaviour
to the caller.

@@

Programs that use resources such as channels are usually written defensively
with the assumption that calling a function will return \emph{exactly once},
either normally or exceptionally. Since effect handlers in Multicore OCaml do
not ensure that all the effects are handled, if the function performs an effect
with no matching handler, then the function \emph{will not return at all}. To
remedy this, when such an effect bubbles up to the top-level, we discontinue the
continuation with an |Unhandled| exception so that the exception handlers may
run and clean up the resources.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases. Programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperfom is implemented as a
primitive to avoid executing code on the resumption path.

@@

      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
      \textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
                        & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
      \textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\caption{Operational semantics of Multicore OCaml effect handlers.}
\end{figure*}

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The value are integer constants $n$, continuations $k$, closure
$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
being raised ($\exnval{l}$). The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
$\fl$ is composed of individual frames $r$, which is one of an argument frame
$\farg{e}{\env}$ with the expression $e$ at the argument position of an
application with its environment $\env$, a function frame $\ffun{v}$ with the
value $v$ at the function position of an application, and frames for evaluating
the arguments of an arithmetic expression.

@@

\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}), reductions
always occur on the top-most fiber in the current stack. \textsc{AdminO}
performs administrative reductions. \textsc{CallO} evaluates an OCaml function
on the current stack. \textsc{ExtCall} captures the behaviour of external
calls, which are evaluated on an empty C stack. \textsc{RetToC} returns a value
to the enclosing C stack. In this case, we have exactly one fiber on the stack,
and this was created in the rule \textsc{Callback}, whose handler has identity
return case alone and the environment is empty. \textsc{RetFib} returns value
from a fiber to the previous one, evaluating the body of the return case.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{Hillerstrom20}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation. To
this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
the top fiber, which we know does not have an effect case (see
\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
where the corresponding effect was performed. This is achieved by appending the
captured continuation to the front of the current continuation.

@@

\subsection{Exceptions}

@@

Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore
OCaml. At the bottom of the stack, we have the |handler_info|, which contains
the pointer to parent fiber, and the closures for the value, exception and
effect cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a context
block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
top-level handler that invokes the exception handler |clos_hexn| after
switching to the parent fiber. This is followed by the |pc| of the code that
invokes the value handler |clos_hval| after switching to the parent stack. The
stack is laid out such that when the computation in the handler returns,
control goes to the value handler.

We have the variable-sized area for the OCaml frames next. Since we want the
fibers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| becomes less than the stack threshold (maintained
in the |Caml_state| table), the stack is said to have overflown. On stack
overflow, we copy the whole fiber to a new area with double the size. In
Multicore OCaml, we introduce stack overflow checks into the function prologue
of OCaml functions. In our evaluation of real world OCaml programs, we observed
that most function calls are to leaf functions with small frame size. In order
to take advantage of this fact, we introduce a small red zone at the top of the
stack. The compiler elides the stack overflow check for leaf functions whose
frame size is less than the size of the red zone.

@@

\subsection{External calls and callbacks}

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the stack pointer in the current fiber, saving the allocation pointer in
the |Caml_state|, updating |rsp| to the top of system stack (maintained in
|Caml_state|), and calling the C function. The actions are reversed when
returning from the external call. For C functions that take arguments on the
stack, the arguments must be copied to the C stack from the OCaml stack where
they are laid out. In the vast majority of external calls, we perform only a
few additional instructions over stock OCaml.

@@

\subsection{Effect handlers}
\label{sec:effimpl}

In order to illustrate the implementation of handling effects, consider the
example presented in Figure~\ref{code:effimpl}. The layout of the program state
as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
performs |E| which is handled in the outer-most handler, and is immediately
resumed. The arrows between the fibers are parent pointers. At position |p1|,
|rsp| is at the top of the fiber |f|.

@@

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |context block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

We have utilized a DWARF verification tool~\cite{Bastian19} with extensions to
handle PIE and additional DWARF primitives to flag possible errors in the hand
written DWARF directives where we have altered the OCaml assembler to implement
effects. In three places additional manual verification was required; two were
where the stack is intentionally switched between fibers and one was where the
return to the parent stack on child stack completion could not track stack
resizes and intermediate effect handler frames.

@@

We consider well-behaved Multicore OCaml programs to resume their continuations
\emph{exactly once}. This can be enforced by installing a finaliser on every
captured continuation that |discontinue|s the continuation during finalisation
and ignores the result:
\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) with _ -> ()) k
\end{lstlisting}
However, Multicore OCaml does not enforce this behaviour in order to avoid
installing a finaliser every time a continuation is captured.

@@

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower}. The outliers (on either ends) were due to the difference in the
allocator and the GC between stock and Multicore OCaml. The biggest impact was
the increase in the code size for OCaml code due to stack overflow checks.
Compared to |stock|, the code size for OCaml code 19\% (30\%) more for |MC| and
|MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is effective
at reducing the code size, but further work is required to bring it closer to
|stock|.

>>>>

\begin{abstract}
Effect handlers have been gathering momentum as a mechanism for modular
	programming with user-defined effects. Effect handlers allow for non-local
	control flow mechanisms such as generators, async/await, lightweight threads
	and coroutines to be composably expressed. We present a design and evaluate a
	full-fledged efficient implementation of effect handlers for OCaml, an
	industrial-strength multi-paradigm programming language. Our implementation
	strives to maintain the backwards compatibility and performance profile of
	existing OCaml code. Retrofitting effect handlers onto OCaml is challenging
	since OCaml does not currently have any non-local control flow mechanisms
	other than exceptions. Our implementation of effect handlers for OCaml: {\em
	(i)} imposes negligible overhead on code that does not use effect handlers;
	{\em (ii)} remains compatible with program analysis tools that inspect the
	stack; and {\em (iii)} is efficient for new code that makes use of effect
	handlers.
\end{abstract}

@@

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

@@

Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
when performed returns the result of division-by-zero. If the denominator of
the division is 0, then we perform the |DivBy0| effect with the numerator value
(line 12) so that the context may decide the return value. We obtain the |iOS|
semantics by wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v (* value case *)
| effect (DivBy0 _) k -> continue k Error (* effect case *)
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

The need to host millions of continuations at the same time rules out the use
of a large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks, and external C functions (which do not have stack
overflow checks) must be performed on a separate system stack. Additionally, we
must generate DWARF stack unwind tables for stacks that may be non-contiguous.
In this work, we develop the compiler and runtime support for implementing
efficient effect handlers for OCaml that satisfy these requirements.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, pushes an exception handler frame that either forwards the
exception to the innermost OCaml exception handler (|raise_exn_c| in
Figure~\ref{fig:stock_stack}) or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
common to use them for \emph{local} control flow.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting the GC. For every call point in the program, the OCaml
compiler emits the size of the frame and the set of all live registers in the
frame containing pointers into the heap. During a GC, the OCaml stack is walked
and the roots are marked, skipping over the C frames.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

@@

OCaml programs that use resources such as channels are usually written defensively
with the assumption that calling a function will return \emph{exactly once},
either normally or exceptionally. Since effect handlers in Multicore OCaml do
not ensure that all the effects are handled, if the function performs an effect
with no matching handler, then the function \emph{will not return at all}. To
remedy this, when such an effect bubbles up to the top-level, we discontinue the
continuation with an |Unhandled| exception so that the exception handlers may
run and clean up the resources.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Eff,Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is
important for backwards compatibility as our goal is to retrofit effect
handlers to a language with large legacy codebases. Programs that do not use
effects remain well-typed, and those that do compose well with those that
don't.

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperfom is implemented as a
primitive to avoid executing code on the resumption path.

@@

      \textsc{Handle}   & (\handle{e}{h}, \env, k, \cstack) & \ostep
                        & \configa{e}{\env}{\ostacka{([],(h,\env)) \kcons k}{\cstack}} \\
      \textsc{ExHn}     & (v, \_, (\ffun{\exnval{l}}::\_,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{x}{e}\} \in h \\
      \textsc{ExFwdC}   & (v, \env, [\ffun{\exnval{l}}::\_,(h,\_)], \cstacka{\fl'}{\ostack}) & \ostep
                        & \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl'}{\ostack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\
      \textsc{ExFwdFib} & (v, \env, (\ffun{\exnval{l}}::\_,(h,\_)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseexn{l}{\_}{\_}\} \notin h \\

      \textsc{EffHn}    & (v, \_, (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons k', \cstack) & \ostep
                        & \configa{e}{\env[r \mapsto k''][x \mapsto v]}{\ostacka{k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{x}{r}{e}\} \in h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\
      \textsc{EffFwd}   & (v, \env', (\ffun{\effval{l}{k}}::\fl,(h,\env)) \kcons (\fl',\hc') \kcons k', \cstack) & \ostep
                        & \configa{v}{\env'}{\ostacka{(\ffun{\effval{l}{k''}}::\fl',\hc') \kcons k'}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } k'' = k ~@~ [(\fl,(h,\env))] \\

      \textsc{EffUnHn}  & (v, \_, [\ffun{\effval{l}{k}}::\fl,(h,\env)], \cstack) & \ostep
                        & \configa{e}{\emptyset}{\ostacka{k ~@~ [(\fl,(h,\env))]}{\cstack}}
                        & \text{if } \{\caseeff{l}{\_}{\_}{\_}\} \notin h \\
                  & & & & \text{and } {e = \throw{\textsf{unhandled}}{0}} \\
      \textsc{Resume}   & (v, \_, (\ffun{k}::\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k', \cstack) & \ostep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k ~@~ ((\fl,\hc) \kcons k')}{\cstack}} \\
    \end{array}
  \end{smathpar}
  \subcaption{OCaml Reductions -- $(\term, \env, k, \cstack) \ostep \config$.}
  \label{sem:ostep}
\end{minipage}
\caption{Operational semantics of Multicore OCaml effect handlers.}
\end{figure*}

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The values are integer constants $n$, continuations $k$, closures
$\clos{x}{e}{\env}$, effects being performed ($\effval{l}{k}$) and exceptions
being raised ($\exnval{l}$). The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the current
continuation $k$ and the C stack $\cstack$ under it. Thus, the program stack is
an alternating sequence of C and OCaml stacks terminating with an empty OCaml
stack $\ostackemp$. The frame list $\fl$ is composed of individual frames $r$,
which is one of an argument frame $\farg{e}{\env}$ with the expression $e$ at
the argument position of an application with its environment $\env$, a function
frame $\ffun{v}$ with the value $v$ at the function position of an application,
and frames for evaluating the arguments of an arithmetic expression.

@@

\textbf{\textit{OCaml reductions.}} In OCaml (Figure~\ref{sem:ostep}),
reductions always occur on the top-most fiber in the current stack.
\textsc{AdminO} performs administrative reductions. \textsc{CallO} evaluates an
OCaml function on the current stack. \textsc{ExtCall} captures the behaviour of
external calls, which are evaluated on an empty C stack with the current OCaml
stack as its tail. \textsc{RetToC} returns a value to the enclosing C stack. In
this case, we have exactly one fiber on the stack, and this was created in the
rule \textsc{Callback}, whose handler has identity return case alone and the
environment is empty. \textsc{RetFib} returns value from a fiber to the
previous one, evaluating the body of the return case.

@@

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation |k''|.
Observe that the captured continuation |k''| includes the current handler.
Intuitively, the handler wraps around captured continuation. This gives
Multicore OCaml effect handlers deep handler semantics~\cite{Hillerstrom20}.
\textsc{EffFwd} forwards the effect to the outer fiber, and extends the
captured continuation |k''| in the process. Recall that we do not capture C
frames as part of a continuation. To this end, \textsc{EffUnHn} models
unhandled effect. If the effect bubbles up to the top fiber, which we know does
not have an effect case (see \textsc{Callback} rule), we raise
\textsc{unhandled} exception at the point where the corresponding effect was
performed. This is achieved by appending the captured continuation to the front
of the current continuation.

@@

\vspace{-2mm}
\subsection{Exceptions}

@@

Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml. At the
bottom of the stack, we have the |handler_info|, which contains the pointer to
parent fiber, and the closures for the value, exception and effect cases. The
closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a context
block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
top-level exception handler frame that forwards exceptions to the parent fiber.
When the exceptions are caught by this handler, the control switches to the
parent stack, and the exception handler closure |clos_hexn| is invoked. This is
followed by the |pc| of the code that returns values to the parent
fiber. This stack is laid out such that when the handled computation returns,
the control switches to the parent fiber and the value handler |clos_hval|
is invoked.

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the stack threshold (maintained in the |Caml_state|
table), the stack is said to have overflown. On stack overflow, we copy the
whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
In our evaluation of real world OCaml programs, we observed that most function
calls are to leaf functions with small frame size. In order to take advantage
of this fact, we introduce a small red zone at the top of the stack. The
compiler elides the stack overflow check for leaf functions whose frame size is
less than the size of the red zone.

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the stack pointer in the current fiber, saving the allocation pointer
value in |r15| in the |Caml_state|, updating |rsp| to the top of system stack
(maintained in |Caml_state|), and calling the C function. The actions are
reversed when returning from the external call. For C functions that take
arguments on the stack, the arguments must be copied to the C stack from the
OCaml stack.

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

In order to illustrate the implementation of handling effects, consider the
example presented in Figure~\ref{code:effimpl}. The layout of the program state
as the program executes is captured in Figure~\ref{fig:fiber_handler}. The code
performs effect |E| which is handled in the outer-most handler, and is
immediately resumed. The arrows between the fibers are parent pointers. At
position |p1|, |rsp| is at the top of the fiber |f|.

@@

The goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers.

We have utilized the DWARF verification tool from~\cite{Bastian19} to flag
possible errors in the hand written DWARF directives where we have altered the
OCaml assembler to implement effect handlers. In three places additional manual
verification was required; two were where the stack is intentionally switched
between fibers and one was where the control returns to the parent fiber which
happened to be resized due to effect handler invocation, which the tool could
not track.

@@

We consider well-behaved Multicore OCaml programs to resume their continuations
\emph{exactly once}. This can be enforced by installing a finaliser on every
captured continuation that |discontinue|s the continuation during finalisation
and ignores the result:
\begin{lstlisting}
Gc.finalise (fun k ->
  try ignore (discontinue k Unwind) with _ -> ()) k
\end{lstlisting}
However, Multicore OCaml does not enforce this behaviour in order to avoid the
costs associated with installing a finaliser every time a continuation is
captured.

@@

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower} than |stock|. The outliers (on either ends) were due to the difference
in the allocator and the GC between stock and Multicore OCaml. The biggest
impact was the increase in the code size for OCaml code due to stack overflow
checks. Compared to |stock|, the code size for OCaml code 19\% (30\%) more for
|MC| and |MC+RedZone32| (|MC+RedZone0|). The result shows that our red zone is
effective at reducing the code size, but further work is required to bring it
closer to |stock|.
